Software Quality Assurance in INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures

### 2019-09-16 paper-jgc.tex
Finishing conclusions & remove what it was called background

%\title{Evolution of Processes in Delivering Reliable Software within the Distributed European Research Infrastructures%\thanks{Grants or other notes
%%about the article that should go on the front page should be
%%placed here. General acknowledgments should be placed at the end of the article.}
%}
\title{Delivering more Reliable Software within the INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
%\subtitle{}

@@

Throughout the last 15 years, Europe has have continuously invested on software development initiatives to provide
research e--Infrastructures with the capability of supporting unified access to large--scale computing and intense 
data analysis facilities. The developed tools have enabled the federation of geographically distributed resource 
providers, with the ultimate aim of supporting the growing needs of the scientific comumunities. According to their
requirements, those research e--Infrastructures have been evolving, exploiting diverse distributed computing 
technologies namely Grid, in the first place, and, eventually, Cloud.

The urging needs of providing stable e--Infrastructures, which were based on novel distributed computing technologies, 
initially drove the efforts spent in software development towards the provision of new functionalities, relegating the 
adoption of software engineering methodologies to a second place. The released software was less tested, resulting in 
a high rate of rollbacks and patches applied to production systems, when compared to the subsequent projects. 
Thereby, the stability and reliability of the e--Infrastructures were often compromised \cite{lingrand-egee}, leading 
to outages in the participant resource centers. Back in the days when Grid middleware was actively developed, 
researchers reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, 
eventually developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.

@@

The increasing prominence of software quality procedures observed in the analysis of the last decade of EU--funded
projects related to the development of distributed computing solutions --as described in Section~\ref{sec:ev}--, is 
pictured in the software project herein introduced: {\sl INDIGO--DataCloud}. As we will thoroughly describe in the next
sections, the ground base laid by these preceding projects and the adoption of practices taken from the prevailing 
software engineering methodologies --in particular the Agile Software Development \cite{agile-manifesto} and the DevOps
culture \cite{zhu}--, were the catalyst to establish a set of Software Quality Assurance (SQA) criteria 
-- as explained in Section~\ref{software-quality-procedures} -- that guided the software lifecycle of the software 
produced.

@@

The fundamental shift brought by the INDIGO--DC SQA procedures was 
to act at the early stages of the software development lifecycle. Thus, instead of validating the software product once 
it was packaged within a pre-release stage, the source code itself is verified for each minor change. This strategy 
allows to improve the sustainability of the software, by checking the compliance with code style standards, and its 
reliability, through the execution of functional and integration tests, ensuring the total absence of regressions. The results
presented in this paper show a divergence between the high rate of bug fixes resolved at development time and the 
significantly lower number of bugs detected by the end users. Consequently, bugs are more effortessly fixed since they 
are early captured and end users interface with more reliable and stable software.

The paper is organized as follows. Section~\ref{sec:b} summarizes the essential context of the research problem.
Section~\ref{sec:ev} details the evolution of research on software reliability in the selected EC--projects. 
Section~\ref{sec:ntsr} provides information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.

%%%%%%%%%%%%%%%%%%%%%%%%

%In this paper, we outline the set of EC--funded projects, ranging from 2001 to nowadays,
%that have raised the awareness and, at the same time, laid the ground work for the establishment of 
%thorough SQA practices to guide the software development lifecycle within the {\sl INDIGO--DataCloud}
%project. 
%%
%%For each project, the challenges, the methodologies and the achievements towards software 
%%reliability are described.
%%During the last 15 years the provision of European e--Infrastructures,
%%supporting unified access to large--scale computing and intense data analysis
%%applications have been driven by the needs of scientific communities as part of their
%%pan--European research collaborations. 
%%One such example is the Worldwide
%%Large Hadron Collider (LHC) Computing Grid infrastructure, driven
%%by the particle physics scientific community, built to satisfy the computing needs
%%of processing, storing and analysis of the very large amount of data produced by the
%%LHC particle accelerator.
%
%A significant number of the innovation projects and technology transfer
%activities presented in Section~\ref{sec:ev}
%were funded by the 5$^{th}$, 6$^{th}$ and 7$^{th}$ Framework Programmes of the European Commission (EC)
%and recently by the Horizon 2020 programme \cite{h2020}. Their main goal was to develop
%software solutions that provide a seamless and transparent access by researchers to
%distributed computing and data resources.
%
%Major developments carried throughout the years were focused on providing new
%functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority.
%The released software was less tested, resulting in a higher rate of
%rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
%reliability of the e--Infrastructures was often compromised \cite{lingrand-egee}, leading to outages
%in the participant resource centers. Back in the days when Grid middleware was actively developed, researchers
%reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, eventually
%developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.
%
%In an effort to improve the research community's experience, it became apparent the need for a better
%balance between the addition of new features and the reliability of the released software. Consequently,
%incoming European funding programmes and open calls stressed the fact of devoting a substantial part of the software lifecycle to the improvement of the quality of the software being delivered. At this point, evolving software engineering practices were considered
%and gradually adopted to define and implement quality procedures, the organization of
%cross--functional teams and deployment of pilot testbeds.
%
%As a result of the analysis the aforementioned projects, 
%one can observe a continuous increase in the prominence and robustness of software
%testing and validation procedures \cite{aiftimiei}. The
%evolution of the software engineering methodologies (creation of the Agile
%Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
%with the parallel advancements in the information and communication technology
%field (such as virtualization, the enabling technology for Cloud computing and
%operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend.
%This evolution was complemented by the emergence of automation and event--response tools.
%
%The former know-how combined with the continuous advancements in sofware engineering 
%have been the catalysts for drawing attention to the relevance of SQA ..

\section{Background}
\label{sec:b}

In this section, we detail the level of knowledge requested in delivering reliable software in order
to understand the effort employed in the EC-funded software projects over time. We divide the literature
into four different domains that intersect this work: 1) software requirements analysis, 2) software
coding,
3) metrics and measurements, 4) testing.

\emph{Software requirements analysis.} They are defined at various stages of a software development process
as a specification of what should be implemented. According to IEEE \cite{radatz}, requirement
analysis is a process for studying user needs to arrive at a definition of system, hardware or software
requirements. Characteristics of requirements that can assess software reliability are consistency and
correctness \cite{sommerville}: the former aims at providing predictable, maintainable and reliable
results, and the latter usually refers to the satisfaction of certain goals.

\emph{Software coding.} According to IEEE \cite{radatz}, coding is the process of expressing a computer
program in a programming language. To meet specific requirements, a program can be developed by using
high-level or low-level programming languages that envelope different characteristics and may affect the
result and quality of software. Furthermore, software is generally written by geographically distributed
teams whose developers can change over time, therefore it is extremely important for source code to be
understandable \cite{tashtoush}. There are various instruments that can be adopted during the software
development process to easily adapt software: programming features that influence the
code readability and maintainability \cite{buse} and standardized code that promotes the comparability of results \cite{baggen}.

\emph{Metrics and measurements.} According to IEEE \cite{radatz}, metrics is a function, with input as
the software data, and output is a value which could decide on how the given attribute affects the software.
Metrics and measurements have an important role in the development process \cite{fenton, capers}. Metrics
are used to improve the ability to control, identify and measure the essential parameters of software
during its development, leading to improve the results of the software product \cite{goodman}. With a correct
usage of metrics, it is possible to determine which parts of the software require to be changed or modified.
In the literature, different types of metrics exist to be able to provide information on various software
aspects, such as complexity \cite{mccabe} and object orientation \cite{chidamber, lorenz}. Often they have
similar definitions, so as a consequence they can be implemented in different ways according to the
interpretation of their definition. This is the reason why it has been always a challenge to select the
proper software metric tools.

\emph{Testing.} According to IEEE \cite{radatz}, testing is a process of operating a system or component
under specified conditions, observing or recording the results, and making an evaluation of some aspect of
the system or component. Testing is an important factor for software quality and reliability improvement. This
activity is quite complicated due to the variety of programming languages, operating systems and hardware
platforms that evolve over time. In the current literature, there are many types of software
testing methods \cite{myers} that aim at finding errors, bugs and faults during the software development
process, possibly in the early stages. Over time, testing metrics have been defined to assess software
quality \cite{athanasiou}: code coverage is the most frequently used metric for testing software reliability
and there exist many tools for dynamic code coverage estimation according to the programming language
\cite{horgan}. A dedicated effort has to be taken into account in order to integrate testing solutions
during the software development process.

\section{Evolution of research on software reliability in EC--projects}
\label{sec:ev}

The EC projects considered in this paper are listed chronologically in Table~\ref{tab:eup}.
Each project paved the way for the following ones, according to a logic evolution in which
the role of the quality and reliability of the software has
consistently increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
committed to, and aware of, software reliability practices.
Table~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Both reliability and sustainability of the {\sl INDIGO--DataCloud} software products 
have been primarily improved by acting at the source code level. According to the 
size of the project, which comprised more than 30 products and 200 GitHub repositories, 
this strategy required a remarkable preliminary ground work. On the one hand, the 
definition of a set of SQA requirements -- compiled in the described SQA criteria 
-- that shall be fulfilled prior to release by the {\sl INDIGO--DataCloud} core 
products. On the other hand, the need of setting up the testing infrastructure and
the associated CI/CD pipelines to check that the changes in the source code matched 
those requirements in an automated fashion. As a complement to the above strategy, a 
final validation step combined the deployment of the pre--released software versions 
in the project's preview testbed, and subsequently, most of the products were tested
in selected production environments through the aid of EGI's staged--rollout process.

The most noticeable metric that demonstrates a reliability improvement comes from the 
fact that the number of software defects uncovered throughout the pre--release phase 
surpassed by a factor of 30 the ones reported by end users in the EGI e-Infrastructure
during the next two years after the first major release (INDIGO-1). Accordingly, the
sustainability of the software was ...

Nevertheless, the individual products of {\sl INDIGO--DataCloud} did not undergo 
a progressive quality improvement throughout the lifespan of the project. Indeed,
unit testing coverage dropped for about 18\% of the total products halfway through
the project. This behavior was aligned with the high-demanding periods of software
development, in particular, the previous months before the second major release 
(INDIGO--2). This fact demonstrates that, even considered as best practice, test 
cases are not commonly written right after, or even before, introducing a new unit
or functionality in the code. High coverage values are generally advisable but they
entail time--consuming work that is often neglected by the code owners. The effort 
spent in testing should be oriented to those sections of the code that are more 
likely to cause negative effects to end users. Consequently, a high coverage value 
does not generally provide a linear benefit with respect to the effort spent -- the
SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
it does not indicate whether the key sections of the code are covered. However, unit
testing is a valuable tool to detect defects at coding time, so whenever the test 
cases accompany the changes done in the code, a high unit testing coverage value is 
a good indicator of reliability in the software.
%(related defect increase?, resulting from the INDIGO-2 release?)

% products not developed from scratch challenged the application of SQA policies??

applying the SQA criteria in the development stage. As each change in 
the source code is validated to match the SQA requirements, defects in the software 
are more likely to be detected and fixed at the earliest stage. Consequently, costs
are reduced, since solving bugs at later stages require greater efforts, and the end
user obtains a much more reliable piece of software, as demonstrated by the low rate
of defects found once it is in production. Automation is of paramount importance to 
guarantee the {\sl INDIGO--DataCloud}'s SQA criteria compliance. Thus, through the
application of DevOps practices, a CI/CD scenario was put in place to cover both the
validation and delivery of the {\sl INDIGO--DataCloud} software products. The CI/CD 
pipelines provide the agility to produce releases frequently enough to reduce the 
impact on production systems, since the software gets progressively updated.

In this regard, a pre--production validation,
through the preview testbeds or resource center's early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

The software lifecycle demands as well a continuous measurement and analysis
process, which allows timely reactions to discover and fix defects throughout the
development, release and maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

Lessons learned show that sustainability is a key pursuit. The achievements, in terms of
processes, knowledge, software and infrastructures, shall be profited from former successful
completed projects. In this regard, the more recent project described in this paper, {\sl INDIGO-DataCloud},
may have set the path for future reference in the European software development research ecosystem.
The SQA criteria, describing the quality requirements that might be considered during the entire
software lifecycle, have been published as open access and periodically reviewed to incorporate
new recommendations and best practices. Successful software products created during the lifetime of the
project are maintained leveraging the same source code repositories and CI infrastructure. A seamless
knowledge transfer among subsequent software development initiatives is key for the good progress of
delivering reliable software.

>>>>

%\title{Evolution of Processes in Delivering Reliable Software within the Distributed European Research Infrastructures%\thanks{Grants or other notes
%%about the article that should go on the front page should be
%%placed here. General acknowledgments should be placed at the end of the article.}
%}
%\title{Delivering more Reliable Software within the INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
\title{Software Quality Assurance in INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
%\subtitle{}

@@

Throughout the last 15 years, the European Commission (EC) has continuously invested on software development initiatives to provide
research e--Infrastructures with the capability of supporting unified access to large--scale computing and intense 
data analysis facilities. The developed tools have enabled the federation of geographically distributed resource 
providers, with the ultimate aim of supporting the growing needs of the scientific comumunities. According to their
requirements, those research e--Infrastructures have been evolving, exploiting diverse distributed computing 
technologies namely Grid, in the first place, and, eventually, Cloud.

The urging needs of providing stable e--Infrastructures, which were based on novel distributed computing technologies, 
initially drove the efforts spent in software development towards the provision of new functionalities, reducing the 
adoption of software engineering methodologies to a marginal player. The released software was less tested, resulting in 
a high rate of rollbacks and patches applied to production systems, when compared to the subsequent projects. 
Thereby, the stability and reliability of the e--Infrastructures were often compromised \cite{lingrand-egee}, leading 
to outages in the participant resource centers. Back in the days when Grid middleware was actively developed, 
researchers reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, 
eventually developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.

@@

The increasing prominence of software quality procedures observed in the analysis of the last decade of EU--funded
projects related to the development of distributed computing solutions --as described in Section~\ref{sec:ev}--, is 
pictured in the software project herein described: {\sl INDIGO--DataCloud}. As we will thoroughly describe in the next
sections, the ground base laid by these preceding projects and the adoption of practices taken from the prevailing 
software engineering methodologies --in particular the Agile Software Development \cite{agile-manifesto} and the DevOps
culture \cite{zhu}--, were the catalyst to establish a set of Software Quality Assurance (SQA) criteria 
-- as explained in Section~\ref{software-quality-procedures} -- that guided the software lifecycle of the software 
produced.

@@

The fundamental shift brought by the {\sl INDIGO--DataCloud} SQA procedures was 
to act at the early stages of the software development lifecycle. Thus, instead of validating the software product once 
it was packaged within a pre-release stage, the source code itself is verified for each minor change. This strategy 
allows to improve the sustainability of the software, by checking the compliance with code style standards, and its 
reliability, through the execution of functional and integration tests, ensuring the total absence of regressions. The results
presented in this paper show a divergence between the high rate of bug fixes resolved at development time and the 
significantly lower number of bugs detected by the end users. Consequently, bugs are more effortessly fixed since they 
are early captured and end users interface with more reliable and stable software.

The paper is organized as follows. %Section~\ref{sec:b} summarizes the essential context of the research problem.
Section~\ref{sec:ev} details the evolution of research on software reliability in the selected EC--projects. 
Section~\ref{sec:ntsr} provides information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.

%\section{Background}
%\label{sec:b}
%In this section, we detail the level of knowledge requested in delivering reliable software in order
%to understand the effort employed in the EC-funded software projects over time. We divide the literature
%into four different domains that intersect this work: 1) software requirements analysis, 2) software
%coding,
%3) metrics and measurements, 4) testing.
%\emph{Software requirements analysis.} They are defined at various stages of a software development process
%as a specification of what should be implemented. According to IEEE \cite{radatz}, requirement
%analysis is a process for studying user needs to arrive at a definition of system, hardware or software
%requirements. Characteristics of requirements that can assess software reliability are consistency and
%correctness \cite{sommerville}: the former aims at providing predictable, maintainable and reliable
%results, and the latter usually refers to the satisfaction of certain goals.
%\emph{Software coding.} According to IEEE \cite{radatz}, coding is the process of expressing a computer
%program in a programming language. To meet specific requirements, a program can be developed by using
%high-level or low-level programming languages that envelope different characteristics and may affect the
%result and quality of software. Furthermore, software is generally written by geographically distributed
%teams whose developers can change over time, therefore it is extremely important for source code to be
%understandable \cite{tashtoush}. There are various instruments that can be adopted during the software
%development process to easily adapt software: programming features that influence the
%code readability and maintainability \cite{buse} and standardized code that promotes the comparability of results \cite{baggen}.
%\emph{Metrics and measurements.} According to IEEE \cite{radatz}, metrics is a function, with input as
%the software data, and output is a value which could decide on how the given attribute affects the software.
%Metrics and measurements have an important role in the development process \cite{fenton, capers}. Metrics
%are used to improve the ability to control, identify and measure the essential parameters of software
%during its development, leading to improve the results of the software product \cite{goodman}. With a correct
%usage of metrics, it is possible to determine which parts of the software require to be changed or modified.
%In the literature, different types of metrics exist to be able to provide information on various software
%aspects, such as complexity \cite{mccabe} and object orientation \cite{chidamber, lorenz}. Often they have
%similar definitions, so as a consequence they can be implemented in different ways according to the
%interpretation of their definition. This is the reason why it has been always a challenge to select the
%proper software metric tools.
%\emph{Testing.} According to IEEE \cite{radatz}, testing is a process of operating a system or component
%under specified conditions, observing or recording the results, and making an evaluation of some aspect of
%the system or component. Testing is an important factor for software quality and reliability improvement. This
%activity is quite complicated due to the variety of programming languages, operating systems and hardware
%platforms that evolve over time. In the current literature, there are many types of software
%testing methods \cite{myers} that aim at finding errors, bugs and faults during the software development
%process, possibly in the early stages. Over time, testing metrics have been defined to assess software
%quality \cite{athanasiou}: code coverage is the most frequently used metric for testing software reliability
%and there exist many tools for dynamic code coverage estimation according to the programming language
%\cite{horgan}. A dedicated effort has to be taken into account in order to integrate testing solutions
%during the software development process.
%\section{Evolution of research on software reliability in EC--projects}
\section{Background work on European e--Infrastructure projects}
\label{sec:ev}

One of the fundamental pillars of the {\sl INDIGO--DataCloud}'s SQA process has been established
by the outcomes resulted from similar EC--funded e--Infrastructure--related projects over
the last 15 years. In particular, in this section we are primarily interested in those projects that consisted in
the collaborative development of software solutions, intended for improving the accessibility and exploitation 
of distributed computing technologies by research communities within the European research
landscape. We will highlight the most representative software engineering
practices adopted within each of these projects to understand how they inspired the SQA process 
definition and implementation of {\sl INDIGO--DataCloud}.

The EC projects considered in this paper are listed chronologically in Table~\ref{tab:eup}.
Each project paved the way for the following ones, according to a logic evolution in which
the quality and reliability of the software solutions have progressively increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
committed to, and aware of, software reliability practices.
Table~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Both reliability and sustainability of the {\sl INDIGO--DataCloud} software products 
have been primarily improved by acting at the source code level. According to the 
size of the project, which comprised more than 30 products and 200 GitHub repositories, 
this strategy required a remarkable preliminary ground work. On the one hand, the 
definition of a set of SQA requirements -- compiled in the described SQA criteria 
-- that ought to be fulfilled by the set of core products before being
released. On the other hand, the setup of a testing infrastructure and
the associated CI/CD pipelines to guarantee that the changes done at the source code
level match
those requirements in an automated fashion. As a final step, the validation of the 
resultant pre--released software versions, firstly tackled through integration testing in
the project's preview testbed, and subsequently complemented by the EGI's 
staged--rollout process, where the brand new versions are adopted by candidate resource
providers within the e-Infrastructure -- accessible for user evaluation --, before being
tagged as production--ready software, and consequently, distributed through the official
EGI channels.

The most noticeable metric that demonstrates a reliability improvement is represented 
by the ratio between the number of software defects uncovered throughout the pre--release
phase of each of the major releases, {\sl INDIGO--1} and {\sl INDIGO--2}, and the ones 
reported by the end users within the EGI e--Infrastructure, where the core products were 
deployed. The number of bugs detected throughout the SQA process here described 
surpassed by a factor of 30 the number of bugs reported by those end users, over a total
of almost 200 bugs in the observed timeframe. Considering that this low rate of bugs -- 
taking into account the amount of software products existing in the project -- 
correspond to the outcomes of the integration testing phase, we infer that the preceding
phase of code analysis -- comprised by the style standard compliance, as well as the unit and
functional test cases execution for each minor change -- had a substantial impact on the robustness
of the resultant software.
% graph with GitHub defects/issues in development
% graph showing types of defects detected in the integration phase

Nevertheless, the individual products of {\sl INDIGO--DataCloud} did not undergo 
a progressive quality improvement throughout its lifespan. Indeed,
unit testing coverage dropped for about 18\% of the total products halfway through
the project. This behavior was aligned with the high-demanding periods of software
development, in particular, the previous months before the second major release 
(INDIGO--2). This fact demonstrates that, even considered as best practice, test 
cases are not commonly written right after, or even before, introducing a new unit
or functionality in the code. High coverage values are generally advisable but they
entail time--consuming work that is often neglected by the code owners. Rather than
requiring a high coverage value, our experience showed that increasing the coverage
of those sections in the code that are more likely to cause negative effects to end
users is not only a more effective approach to improve software reliability, but it 
also keeps developers motivated in the tedious task of test coding, as it brings more 
noticeable benefits for the end user.
% ?? (related defect increase?, resulting from the INDIGO-2 release?)

%The effort 
%spent in testing should be oriented to those sections of the code that are more 
%likely to cause negative effects to end users. Consequently, a high coverage value 
%does not generally provide a linear benefit with respect to the effort spent -- the
%SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
%it does not indicate whether the key sections of the code are covered. However, unit
%testing is a valuable tool to detect defects at coding time, so whenever the test 
%cases accompany the changes done in the code, a high unit testing coverage value is 
%a good indicator of reliability in the software.

All along the project we have observed that source code review is the cornerstone to
improve the effectiveness of the unit and functional -- when applicable -- test cases. 
{\sl INDIGO--DataCloud} source code repositories are protected against direct pushes 
so that the code review stage always takes place. Code reviewers analyse the content
of each change in the source code and the suitability -- according to the SQA criteria -- of the
associated tests, being the last checkpoint before the change is added into the
production branch. Hence the vital importance of providing the tests alongside the
code for every given change.
% ?? figure with code reviews or sth about PRs, external participation
%Functional quality, on the other hand, was assessed within the {\sl INDIGO--DataCloud} 
%software lifecycle through the per--change execution of the functional tests provided 
%by the developers and the ultimate stage of integration testing in the aforementioned 
%preview testbed. As we have seen, the latter testing strategy was entirely performed 
%in a manual way, and appeared as a helpful tool to spot defects after exposing the 
%deployed software to interactions with external components. Additionally, the functional 
%tests, were primarily useful to check that the software met the design specifications, 
%and most importantly, that previously working functionalities are preserved when new 
%features are added. As mentioned, these type of testing is meant to be continuously 
%executed in an automated manner using the CI infrastructure. 
However, by the time of
{\sl INDIGO--1} only 30\% of the total of products provided automated functional tests,
reaching almost 60\% in {\sl INDIGO--2}. Although test reports were asked in those cases
where automated functional tests were not provided, the fulfillment of 
their fundamental functional requirements was tougher to track, resulting in a non--viable
solution for assessing minor changes, especially when analysing the regressions. The major issues observed that hindered developers from the 
implementation of automated functional tests were releated to the unavailability
-- at the time -- of libraries and frameworks for tackling specific tests -- such as 
testing graphical interfaces --, or in some cases, the unawareness about their existence.
We infer from this experience that automated functional tests should be a goal in modern
software development as they complement the structural quality checks by adding the
validation of the software's identified functionalities, key in the operational impact 
and user acceptance. 
% figure with % of products with functional tests
% ?? products not developed from scratch challenged the application of SQA policies??

The application of the formerly described SQA process was not only beneficial for the
core products' quality and reliability, but it has proven to be equally effective for
those applications uniquely supported by research communities with little background 
on software engineering. This fact was demonstrated by the elaboration of an ad--hoc validation pipeline for one
of the use cases supported in the project. The CI/CD pipeline extended the source code testing and delivery
-- already existing in the pipelines for the core products -- with a basic system--like testing. 
For the latter, the
pre--packaged application was tested with a pre--defined set of inputs and expected
outputs. Only if successfully validated, the source code could be merged into production 
and the resultant pre--packaged application would be delivered. When this pipeline was
integrated in the usual workflow of the application development, it was reported to have constituted a real 
breakthrough, allowing a faster development and more reliable software, as it was programmatically tested
before being distributed. Consequently, the 
benefits of a continuous SQA improvement process are not only suitable for experienced
software developers, but also to any type of software practitioner, such as the numerous ones
commonly found within the research communities.

The {\sl INDIGO--DataCloud}'s SQA process has persisted once the project concluded, being 
extensively used by the supported software products. The SQA criteria is actively
maintained to include new recommendations and best practices, not only targeted for
experts but for any enthusiast on research software. Hence, they have established the foundations for 
knowledge transfer capabilities, which we foresee it will be convenient for the rapid adoption of 
SQA practices in the subsequent software development initiatives and projects for e--Infrastructure 
building and management, taking a step forward in the overarching goal
of delivering reliable software in research.
